---
title: "20221112_Stock_Portfolio_Analysis"
author: "Felipe Tufaile, Vinicius de Camargo, Helena Funari, Rodrigo Zamengo"
date: "11/12/2022"
output: html_document
---


### Summary

Text text text


```{r configs, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, include = TRUE, message = FALSE)
```

### Loading libraries

```{r libraries}

## Set working directory
setwd("~/Insper/Commodity_Export_Forecast")

library(dplyr)
library(tidyverse)
library(ggthemes)
library(fpp3)
library(forecast)
library(gridExtra)
library(ggpubr)
library(tseries)
library(prophet)
library(rugarch)
library(yfR)
```


```{r reading_stocks, out.width="100%"}

## Defining the start date of the time series
start_date <- '2021-01-01'

## Defining the stocks that will be considered for this analysis
tickers <- c("HGRE11.SA", "MGLU3.SA", "VALE3.SA", "PETR4.SA", "B3SA3.SA", "HASH11.SA")

## Reading the log-return of each chosen stock
#df_log_returns <- yfR::yf_get(
#  tickers=tickers,
#  first_date = start_date,
#  type_return = "log",
#  freq_data = "daily", 
#  do_complete_data = TRUE
#)

setwd("~/Insper/Stock_Portfolio_Analysis")
df_log_returns <- read_csv("20221113_stocks.csv") %>%
    mutate(ret_closing_prices = as.numeric(ret_closing_prices))

## Converting the dataframe into a time series object using tsibble
df_log_returns_tsibble <- df_log_returns %>% 
  as_tsibble(key = ticker, index = ref_date, regular = FALSE)

## Plotting the closing price time series
df_log_returns_tsibble %>%
  autoplot(price_close, colour = "black") +
  facet_wrap(~ticker, scales = "free_y", ncol = 1) +
  ggtitle("Closing price for selected stock and ETFs", "Data ranging from Jan-2021 to Nov-22") +
  labs(x="Reference Date", y = "Closing Price (R$)")
```


```{r plotting_close_price_return, out.width="100%"}
## Plotting the closing price return time series
df_log_returns_tsibble %>%
  autoplot(ret_closing_prices, colour = "black") +
  facet_wrap(~ticker, scales = "fixed", ncol = 1) +
  ggtitle("Log Return of Closing Price for selected stock and ETFs", "Data ranging from Jan-2021 to Nov-22") +
  labs(x="Reference Date", y = "Log Return of Closing Price")
```

```{r filtering_domain}
## Calculating the first not null day: when calculating log returns the first day becomes null
first_date <- df_log_returns_tsibble %>%
  dplyr::group_by(ticker) %>% 
  dplyr::filter(ref_date == min(ref_date)) %>% 
  dplyr::ungroup() %>% 
  with(max(ref_date))

## Filtering out all data before "first_date"
df_log_returns_cln_tsibble <- df_log_returns_tsibble %>%
    dplyr::filter(ref_date > first_date)
```


# Descritivas bacanas

- ACF/PACF dos retornos
- visualizar os retornos ao quadrado
- ACF/PACF dos retornos ao quadrado

```{r plotting_acf, out.width=100%}
df_log_returns_cln_tsibble %>%
  ACF(ret_closing_prices) %>%
  autoplot()
```



```{r plotting_pacf, out.width=100%}
df_log_returns_cln_tsibble %>%
  PACF(ret_closing_prices) %>%
  autoplot()
```


```{r calculating_return_squared, out.width=100%}
df_log_returns_cln_tsibble <- df_log_returns_cln_tsibble %>%
  dplyr::mutate(ret2 = ret_closing_prices^2) 

df_log_returns_cln_tsibble %>%
  autoplot(ret2, colour = "black") +
  facet_wrap(~ticker, ncol = 1, scales = "fixed")
```


```{r plotting_acf_log_return_sqrd, out.width=100%}
df_log_returns_cln_tsibble %>%
  ACF(ret2) %>% 
  autoplot()
```

```{r plotting_pacf_log_return_sqrd, out.width=100%}
df_log_returns_cln_tsibble %>% 
  PACF(ret2) %>% 
  autoplot()
```



```{r}
df_log_returns_cln_tsibble %>%
  group_by(ticker) %>%
  summarize(gg=list(ggplot(cur_data(), aes(sample=ret_closing_prices)) + geom_qq() + geom_qq_line() + labs(title=cur_group()))) %>% 
  dplyr::pull(gg) %>% 
  patchwork::wrap_plots()
```


```{r}
df_log_returns_cln_tsibble %>%
  group_by(ticker) %>%
  summarize(gg=list(ggplot(cur_data(), aes(sample=ret_closing_prices)) + 
                      geom_qq(distribution=qt, dparams=list(df=3)) + 
                      geom_qq_line(distribution=qt, dparams=list(df=3)) + 
                      labs(title=cur_group()))) %>% 
  dplyr::pull(gg) %>% 
  patchwork::wrap_plots()
```

## Ajustando modelos garch


```{r}
## Adjusting a garch model
garch_individual <- function(parms, ret, prog = NULL) {
  if (!is.null(prog)) prog()
  # daria para adicionar mais hiperparametros!!!
  garch_model = ugarchspec(
    variance.model = list(
      model = "fGARCH",
      submodel = "GARCH",
      garchOrder = c(parms$m, parms$n)
    ),
    mean.model = list(
      armaOrder = c(parms$p, parms$q),
      include.mean = TRUE
    ),
    distribution.model = parms$dist
  )
  # as vezes ele nao converge
  suppressWarnings({
    fit <- ugarchfit(garch_model, data = ret)
  })
  fit
}
```


Função para ajustar uma grid de garchs e pegar as informações

```{r}
melhor_garch <- function(ativo, p = 0:3, q = 0:3, m = 0:2, n = 0:2, dist = "std") {
 
  # faz uma grid de hiperparâmetros 
  grid <- expand_grid(p = p, q = q, m = m, n = n, dist = dist) |> 
    tibble::rownames_to_column("id")
  
  # pega o retorno do ativo
  ret <- da_train |> 
    dplyr::filter(ticker == ativo) |> 
    pull(ret_closing_prices)
  
  usethis::ui_info("Ajustando modelos para {ativo}...")
  
  progressr::with_progress({
    prog <- progressr::progressor(nrow(grid))
    modelos <- grid |> 
      group_split(id) |> 
      purrr::map(garch_individual, ret = ret, prog)
  })
  
  safe_info <- purrr::possibly(infocriteria, tibble::tibble())
  
  suppressWarnings({
    informacao <- purrr::map(modelos, safe_info) |> 
      purrr::map(tibble::as_tibble, rownames = "criteria") |> 
      dplyr::bind_rows(.id = "id")
  })
  
  melhores <- informacao |> 
    dplyr::inner_join(grid, "id") |> 
    tidyr::pivot_wider(names_from = criteria, values_from = V1) |> 
    janitor::clean_names() |> 
    dplyr::arrange(akaike)
  
  usethis::ui_info(c(
    "Melhor modelo:",
    "p <- {melhores$p[1]}",
    "q <- {melhores$q[1]}",
    "m <- {melhores$m[1]}",
    "n <- {melhores$n[1]}"
  ))
  
  melhores
  
}

```

Rodando as funções

```{r}
#| eval: false
melhores_por_ativo <- ativos |> 
  purrr::set_names() |> 
  purrr::map(melhor_garch) |> 
  dplyr::bind_rows(.id = "ticker")
```

```{r}
#| echo: false
melhores_por_ativo <- readr::read_rds("melhores_por_ativo.rds")
```


## Prever volatilidade um passo à frente

Função que ajusta o modelo e faz as previsões

```{r}
prever_volatilidade <- function(parms, n_steps = 5) {
  
  ret <- da_train |> 
    dplyr::filter(ticker == parms$ticker) |> 
    pull(ret_closing_prices)
  
  garch_model = ugarchspec(
    variance.model = list(
      model = "fGARCH",
      submodel = "GARCH",
      garchOrder = c(parms$m, parms$n)
    ),
    mean.model = list(
      armaOrder = c(parms$p, parms$q),
      include.mean = TRUE
    ),
    distribution.model = parms$dist
  )
  
  fit <- ugarchfit(garch_model, data = ret, out.sample = n_steps - 1)
  
  # browser()
  
  if (parms$dist == "std") {
    shape <- as.numeric(fit@fit$coef["shape"])
  } else {
    shape <- NA_real_
  }
  
  forecasts <- ugarchforecast(fit, n.ahead = n_steps)@forecast
  tibble::tibble(
    ticker = parms$ticker,
    serie = as.numeric(forecasts$seriesFor),
    volatilidade = as.numeric(forecasts$sigmaFor),
    shape = shape
  )
}
```

Ajustando modelos finais e prevendo volatilidade futura

```{r}
parametros_melhores <- melhores_por_ativo |> 
  group_by(ticker) |> 
  slice_head(n = 1) |> 
  ungroup()

vol_futuro <- parametros_melhores |> 
  group_split(ticker) |> 
  purrr::map(prever_volatilidade, n_steps = 1) |> 
  dplyr::bind_rows()

vol_futuro
```
```{r}
vol_futuro[1,3]
```


## Comparar volatilidades entre os retornos selecionados

#Calcular os VaR
Considerando: 99% de confiança
VaR = 2,33 * raiz de h * valor investido


```{r}
#B3SA3.SA

vars <- c(VaR_B3SA3 = 2.33 * vol_futuro$volatilidade[1] * 10000,

VaR_HGRE11 = 2.33 * vol_futuro$volatilidade[2] * 10000,

VaR_MGLU3 = 2.33 * vol_futuro$volatilidade[3] * 10000,

VaR_PETR4 = 2.33 * vol_futuro$volatilidade[4] * 10000,

VaR_VALE3 = 2.33 * vol_futuro$volatilidade[5] * 10000)

tickers <- c('B3SA3', 'HGRE11', 'MGLU3', 'PETR4', 'VALE3')

vars <- tibble(tickers = tickers, VaRs = vars)


arrange(vars, desc(VaRs))



```


...

## Montagem de portifolio

Reproduzindo código daqui:

<https://www.codingfinance.com/post/2018-05-31-portfolio-opt-in-r/>

Versão em python

<https://www.codingfinance.com/post/2018-05-31-portfolio-opt-in-python/>

```{r}
da_wide <- da_train |> 
  dplyr::select(ref_date, name = ticker, value = ret_closing_prices) |> 
  tidyr::pivot_wider()

da_xts <- da_wide |> 
  timetk::tk_xts(select = -ref_date, date_var = ref_date)
```


```{r}
mean_ret <- colMeans(da_xts, na.rm = TRUE)
print(round(mean_ret, 5))
```

Next we will calculate the covariance matrix for all these stocks. We will NOT annualize it by multiplying by 252.

```{r}
cov_mat <- cov(da_xts, use = "complete.obs")
print(round(cov_mat,6))
```

Before we apply our methods to thousands of random portfolio, let us demonstrate the steps on a single portfolio.

To calculate the portfolio returns and risk (standard deviation) we will us need

- Mean assets returns
- Portfolio weights
- Covariance matrix of all assets
- Random weights

```{r}
set.seed(2)
# Calculate the random weights
wts <- runif(n = length(ativos))

(wts <- wts/sum(wts))

# Calculate the portfolio returns
(port_returns <- sum(wts * mean_ret))

# Calculate the portfolio risk
(port_risk <- sqrt(t(wts) %*% (cov_mat %*% wts)))

# Calculate the Sharpe Ratio
(sharpe_ratio <- port_returns/port_risk)
```

We have everything we need to perform our optimization. All we need now is to run this code on 5000 random portfolios. For that we will use a for loop. 


~Before we do that, we need to create empty vectors and matrix for storing our values.~

```{r}

sim_returns <- function(i) {
  wts <- runif(length(ativos))
  wts <- wts / sum(wts)
  port_ret <- sum(wts * mean_ret)
  port_sd <- as.numeric(sqrt(t(wts) %*% (cov_mat %*% wts)))
  sr <- port_ret / port_sd
  
  wts |> 
    purrr::set_names(tickers) |> 
    tibble::enframe() |> 
    tidyr::pivot_wider() |> 
    dplyr::mutate(
      return = port_ret,
      risk = port_sd,
      sharpe = sr
    )
}

portfolio_values <- purrr::map(1:5000, sim_returns) |> 
  bind_rows(.id = "run")

min_var <- portfolio_values[which.min(portfolio_values$risk),]
max_sr <- portfolio_values[which.max(portfolio_values$sharpe),]

```

Lets plot the weights of each portfolio. First with the minimum variance portfolio.

```{r}

min_var |> 
  pivot_longer(2:6) |> 
  mutate(name = forcats::fct_reorder(name, value)) |> 
  ggplot(aes(name, value)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Asset",
    y = "Weight",
    title = "Minimum variance portfolio weights"
  )

```

```{r}
max_sr |> 
  pivot_longer(2:6) |> 
  mutate(name = forcats::fct_reorder(name, value)) |> 
  ggplot(aes(name, value)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Asset",
    y = "Weight",
    title = "Tangency portfolio weights"
  )
```

```{r}
portfolio_values |> 
  ggplot(aes(x = risk, y = return, color = sharpe)) +
  geom_point() +
  theme_classic() +
  scale_y_continuous(labels = scales::percent) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    x = 'Risk',
    y = 'Returns',
    title = "Portfolio Optimization & Efficient Frontier"
  ) +
  geom_point(
    aes(x = risk, y = return), 
    data = min_var, 
    color = 'red'
  ) +
  geom_point(
    aes(x = risk, y = return), 
    data = max_sr, 
    color = 'red'
  )
```

## VaR do portfolio

```{r}
pesos_finais <- min_var |> 
  dplyr::select(2:6) |> 
  as.numeric()


rt_final <- mean(vol_futuro$serie * pesos_finais)
st_dev_final <- sqrt(pesos_finais %*% cov_mat %*% pesos_finais)
nu <- min(vol_futuro$shape)
valor_t <- qt(.95, nu)

(VaR <- rt_final + valor_t * st_dev_final / sqrt(nu/(nu-2)))
```

## CAPM

```{r}
portfolio_returns <- da_train |> 
  tidyquant::tq_portfolio(
    ticker, 
    ret_closing_prices, 
    weights = pesos_finais,
    col_rename = "portfolio"
  )

market_returns <- yfR::yf_get(
  "^BVSP",
  first_date = data_corte + 1,
  type_return = "log",
  freq_data = "daily", 
  do_complete_data = TRUE
) |> 
  dplyr::select(ref_date, ibov = ret_closing_prices)

all_returns <- market_returns |> 
  dplyr::inner_join(portfolio_returns, "ref_date") |> 
  tidyr::drop_na()

(beta_geral <- with(all_returns, cov(portfolio, ibov) / var(ibov)))

calcular_beta <- function(ativo) {
  da_train |> 
    dplyr::filter(ticker == ativo) |> 
    dplyr::inner_join(market_returns, "ref_date") |> 
    tidyr::drop_na() |> 
    with(cov(ret_closing_prices, ibov) / var(ibov))
}

betas <- purrr::map_dbl(ativos, calcular_beta) |> 
  purrr::set_names(ativos)

sum(betas * pesos_finais)
beta_geral

```

```{r}

capm_lm_tudo <- lm(portfolio ~ ibov, data = all_returns) |> 
  broom::tidy() |> 
  dplyr::filter(term == "ibov") |> 
  with(estimate)

capm_lm_individual <- purrr::map_dbl(ativos, \(ativo) {
  da_model <- da_train |> 
    dplyr::filter(ticker == ativo) |> 
    dplyr::inner_join(market_returns, "ref_date")
  lm(ret_closing_prices ~ ibov, data = da_model) |> 
    broom::tidy() |> 
    dplyr::filter(term == "ibov") |> 
    dplyr::pull(estimate)
}) |> 
  purrr::set_names(ativos)


capm_lm_tudo

capm_lm_individual
```